A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-08-29 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAdditiveProbability: Replace Ferrie with Zachos, add Kolmogorov

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories in terms of probabilistic and/or entropic concepts \cite{plavala2023,ariano2017,caticha2019,hall2016,wilce2025}, [CITE - GPT, Ariel, Reginatto/Hall...] which stems from the underlying aim to characterize physical theories from an operational viewpoint. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics\cite{chsh1969}.[CITE] Quasiprobabilistic approaches \cite{ferrie2010,mandel1995,monchietti2025} [CITE] are also used in quantum mechanics, but it is not completely clear how exactly they relate to the standard one.

@@

We will first review how measure theory is used in classical probability [cite] and classical statistical mechanics \cite{Peliti2011}. We will connect this to Choquet theory \cite{phelps2001}, which aims to represent each point of a convex subset as a probability measure over its extreme points (i.e. each ensemble can be represented by a probability measure over pure states). While such representation is always possible, at least in the compact case, it is not unique. Uniqueness requires the space to be a Choquet simplex, the generalization of the finite-dimensional simplex, and represents the classical case. The key insight, then, is that an ensemble space is classical if and only if each ensemble admits a unique decomposition in terms of pure states, it is a Choquet simplex.

>>>>

In the past few decades, there has been a growing interest in developing approaches that generalize the notion of states and physical theories in terms of probabilistic and/or entropic concepts \cite{plavala2023,ariano2017,caticha2019,hall2016,wilce2025}, [CITE - GPT, Ariel, Reginatto/Hall...] which stems from the underlying aim to characterize physical theories from an operational viewpoint. One basic insight is that statistical/probabilistic concepts often provide a suitable basis for this generalization, and it is in this setting where most of these approaches operate. A key problem, then, is what notion of probability should be used given that the standard measure theoretic Kolmogorovian approach does not generally work in quantum mechanics\cite{chsh1969,fine1982}.[CITE] Quasiprobabilistic approaches \cite{zachos2005,mandel1995,monchietti2025} [CITE] are also used in quantum mechanics, but it is not completely clear how exactly they relate to the standard one.

@@

We will first review how measure theory is used in classical probability \cite{kolmogorov2018} and classical statistical mechanics \cite{Peliti2011}. We will connect this to Choquet theory \cite{phelps2001}, which aims to represent each point of a convex subset as a probability measure over its extreme points (i.e. each ensemble can be represented by a probability measure over pure states). While such representation is always possible, at least in the compact case, it is not unique. Uniqueness requires the space to be a Choquet simplex, the generalization of the finite-dimensional simplex, and represents the classical case. The key insight, then, is that an ensemble space is classical if and only if each ensemble admits a unique decomposition in terms of pure states, it is a Choquet simplex.
